Psychological Needs During Intergroup Contact

### 2021-11-23 Methods-and-Results.tex
unnecessary comments removed
wording issues fix

\faQuestionCircle~Unsure about order of study one and two. Currently
chronological. But could also be: (1) main model works in larger student
sample, (2) works in economic migrant sample, (3) test full thing in
medical sample.

@@

We find that in the highest multilevel models, the fulfillment of core
situational needs during outgroup contacts was associated with more
positive outgroup attitudes (random slopes model; \textit{b} = 0.17,
\textit{t}(365) = 2.93, \textit{p} = 0.004, \textit{95\%CI}{[}0.06,
0.29{]}) and also predicted higher perceived interaction quality (random
intercept model; \textit{b} = 0.42, \textit{t}(365) = 8.07, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.32, 0.52{]}). However, if we
consider the influences of core need fulfillment and interaction quality
on outgroup attitudes jointly, we find that virtually all variance is
explained by perceived interaction quality (random slopes mode;
\textit{b} = 0.23, \textit{t}(364) = 4.22, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.33{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.04,
\textit{t}(364) = 0.64, \textit{p} = 0.523, \textit{95\%CI}{[}-0.08,
0.15{]}). We thus find support for our hypotheses and can conclude that
in this data set the fulfillment of core situational needs had a
significant influence on outgroup attitudes and that this effect is
likely explained by its effect through perceived interaction quality.

@@

We additionally used a multilevel regression to check whether having an
interaction with an outgroup member had a situational (i.e.,
contemporaneous) effect within the participants. We find that having an
outgroup interaction is indeed associated with significantly more
positive outgroup attitudes within the participants (random slopes
model; \textit{b} = 2.99, \textit{t}(4850) = 3.80, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.45, 4.54{]}), even after
controlling for having an interaction with a non-Dutch (which did not
relate to outgroup attitudes independently; For full results see Online
Supplementary Materials B). Thus, in our second dataset we find mixed
results, outgroup contacts show a positive effect on outgroup attitudes
in the moment and on average between participants. However, we find that
this effect does not depend on the average perceived interaction
quality. However, this might be due to the aggregation process and the
following analyses will focus in on the role of the perceived
interaction quality during an outgroup interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). However, if we consider the influences
of core need fulfillment and interaction quality on outgroup attitudes
jointly, we find that virtually all variance is explained by perceived
interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}). These results are consistent with the results in study one and
we thus find support for our hypotheses and can conclude that in this
data set the fulfillment of core situational needs had a significant
influence on outgroup attitudes and that this effect is likely explained
by its effect through perceived interaction quality.

@@

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 108) = 22.90,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.08, \textit{t}(823) = 2.95, \textit{p} = 0.003,
\textit{95\%CI}{[}0.03, 0.13{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all of the three fundamental needs. However,
different from the first study, in this larger study relatedness
(\textit{b} = 0.07, \textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) predicted positive outgroup attitudes. For full results see
Online Supplementary Information B.

@@

We sequentially tested whether the fulfillment of Allport's Contact
Conditions was (1) related to more positive outgroup attitudes, (2)
higher perceived interaction quality, and (3) whether the variance
explained by Allport's Conditions is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of Allport's Conditions during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.22, \textit{t}(1601) = 5.86,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.15, 0.29{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.65, \textit{t}(1605) = 11.82, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.54, 0.76{]}). However, if we consider the
influences of Allport's Conditions and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.25, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.23{]}) and the unique variance explained by
Allport's Conditions was less than half of its original effect size
(\textit{b} = 0.09, \textit{t}(1600) = 2.89, \textit{p} = 0.004,
\textit{95\%CI}{[}0.03, 0.15{]}). These results indicate that in this
dataset the the fulfillment of Allport's conditions had a significant
influence on outgroup attitudes and this effect is likely in parts
explained by its effect through perceived interaction quality.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core need fulfillment during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.19, \textit{t}(1601) = 5.29,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.12, 0.27{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.43, \textit{t}(1605) = 8.35, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.33, 0.54{]}). However, if we consider the influences
of core need fulfillment and interaction quality on outgroup attitudes
jointly, we find that perceived interaction quality is a substantially
stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.86, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.22{]}) and the unique variance explained by
core need fulfillment was roughly half of its original effect size
(\textit{b} = 0.11, \textit{t}(1600) = 3.50, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.05, 0.17{]}). These results indicate that in
this dataset the the fulfillment of core situational needs had a
significant influence on outgroup attitudes and this effect is likely in
parts explained by its effect through perceived interaction quality.

@@

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 70) = 100.20,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.15, \textit{t}(1598) = 3.88, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.07, 0.22{]}). When compared to the model with
only the SDT needs, the core need fulfillment flexibly takes on some of
the explained variance of all of the three fundamental needs. However,
similar to the previous study, in this large sample relatedness
(\textit{b} = 0.05, \textit{t}(1598) = 3.31, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.02, 0.09{]}), competence (\textit{b} = 0.06,
\textit{t}(1598) = 2.73, \textit{p} = 0.006, \textit{95\%CI}{[}0.02,
0.10{]}) and autonomy (\textit{b} = 0.04, \textit{t}(1598) = 2.12,
\textit{p} = 0.034, \textit{95\%CI}{[}-0.01, 0.08{]}) each predicted
positive outgroup attitudes independently. However, the regression
coefficient is three times as large for the core need fulfillment (with
all scaling being equal). For full results see Online Supplementary
Information B.

>>>>

@@

We find that in the highest multilevel models, the fulfillment of core
situational needs during outgroup contacts was associated with more
positive outgroup attitudes (random slopes model; \textit{b} = 0.17,
\textit{t}(365) = 2.93, \textit{p} = 0.004, \textit{95\%CI}{[}0.06,
0.29{]}) and also predicted higher perceived interaction quality (random
intercept model; \textit{b} = 0.42, \textit{t}(365) = 8.07, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.32, 0.52{]}). Moreover, when we
consider the influences of core need fulfillment and interaction quality
on outgroup attitudes jointly, we find that virtually all variance is
explained by perceived interaction quality (random slopes mode;
\textit{b} = 0.23, \textit{t}(364) = 4.22, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.33{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.04,
\textit{t}(364) = 0.64, \textit{p} = 0.523, \textit{95\%CI}{[}-0.08,
0.15{]}). We thus find support for our hypotheses and can conclude that
in this data set the fulfillment of core situational needs had a
significant influence on outgroup attitudes and that this effect is
likely explained by its effect through perceived interaction quality.

@@

We additionally used a multilevel regression to check whether having an
interaction with an outgroup member had a situational (i.e.,
contemporaneous) effect within the participants. We find that having an
outgroup interaction is indeed associated with significantly more
positive outgroup attitudes within the participants (random slopes
model; \textit{b} = 2.99, \textit{t}(4850) = 3.80, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.45, 4.54{]}), even after
controlling for having an interaction with a non-Dutch (which did not
relate to outgroup attitudes independently; For full results see Online
Supplementary Materials B). Thus, in our second dataset we find mixed
results, outgroup contacts show a positive effect on outgroup attitudes
in the moment and on average between participants. However, we find that
this effect does not depend on the average perceived interaction
quality. This unexpected results, might be due to the aggregation
process and the following analyses will focus in on the role of the
perceived interaction quality during an outgroup interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). Additionally, if we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that virtually all variance is explained by
perceived interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}). These results are consistent with the results in study one and
we thus find support for our hypotheses and can conclude that in this
data set the fulfillment of core situational needs had a significant
influence on outgroup attitudes and that this effect is likely explained
by its effect through perceived interaction quality.

@@

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 108) = 22.90,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.08, \textit{t}(823) = 2.95, \textit{p} = 0.003,
\textit{95\%CI}{[}0.03, 0.13{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all of the three fundamental needs. However,
different from the first study, in this larger sample relatedness
(\textit{b} = 0.07, \textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) also predicted positive outgroup attitudes. For full results
see Online Supplementary Information B.

@@

We sequentially tested whether the fulfillment of Allport's Contact
Conditions was (1) related to more positive outgroup attitudes, (2)
higher perceived interaction quality, and (3) whether the variance
explained by Allport's Conditions is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of Allport's Conditions during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.22, \textit{t}(1601) = 5.86,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.15, 0.29{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.65, \textit{t}(1605) = 11.82, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.54, 0.76{]}). Moreover, when we consider the
influences of Allport's Conditions and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.25, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.23{]}) and the unique variance explained by
Allport's Conditions was less than half of its original effect size
(\textit{b} = 0.09, \textit{t}(1600) = 2.89, \textit{p} = 0.004,
\textit{95\%CI}{[}0.03, 0.15{]}). These results indicate that in this
dataset the the fulfillment of Allport's conditions had a significant
influence on outgroup attitudes and this effect is likely in parts
explained by its effect through perceived interaction quality.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core need fulfillment during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.19, \textit{t}(1601) = 5.29,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.12, 0.27{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.43, \textit{t}(1605) = 8.35, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.33, 0.54{]}). Additionally, once we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.86, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.22{]}) and the unique variance explained by
core need fulfillment was roughly half of its original effect size
(\textit{b} = 0.11, \textit{t}(1600) = 3.50, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.05, 0.17{]}). These results indicate that in
this dataset the the fulfillment of core situational needs had a
significant influence on outgroup attitudes and this effect is likely in
parts explained by its effect through perceived interaction quality.

@@

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 70) = 100.20,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.15, \textit{t}(1598) = 3.88, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.07, 0.22{]}). When compared to the model with
only the SDT needs, the core need fulfillment flexibly takes on some of
the explained variance of all of the three fundamental needs. However,
similar to the previous study, in this large sample relatedness
(\textit{b} = 0.05, \textit{t}(1598) = 3.31, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.02, 0.09{]}), competence (\textit{b} = 0.06,
\textit{t}(1598) = 2.73, \textit{p} = 0.006, \textit{95\%CI}{[}0.02,
0.10{]}) and autonomy (\textit{b} = 0.04, \textit{t}(1598) = 2.12,
\textit{p} = 0.034, \textit{95\%CI}{[}-0.01, 0.08{]}) each also
predicted positive outgroup attitudes independently. However, the
regression coefficient is three times as large for the core need
fulfillment (with all scaling being equal). For full results see Online
Supplementary Information B.
